Curve counting in genus one: elliptic singularities {\it \&} relative geometry

### 2019-06-28 EllipticSingularitiesSplitting.tex
put back in a bit more discussion of Etilde

\subsubsection{Type $B$}
Now suppose $\Delta$ is of type $B$. Note that in this case, it is impossible for the core to be contracted. Hence $\Ecal$ is disjoint from the blown-up locus, and the map $\Dcal \to \Ecal$ is an isomorphism. We thus obtain the following description, entirely in terms of genus zero data:
\begin{lemma} $\Dcal$ admits a finite and surjective splitting morphism:
\begin{equation*} \Dcal \xrightarrow{\rho} \left(\ol\Mcal_{0,\alpha^{(1)}\cup(m_1,m_2)}(\PP^N|H,d_1)\times\prod_{i=3}^r \ol\Mcal_{0,\alpha^{(i)}\cup(m_i)}(\PP^N|H,d_i)\right) \times_{H^r} \ol\Mcal^{\leftrightarrow}_{0,\alpha^{(0)}\cup(-m_1,\ldots,-m_r)}(P|H_0+H_\infty,d_0).\end{equation*}\end{lemma}

@@

The combinatorics of the flattening can be described explicitly. Consider the map on tropicalizations
\[
\widetilde{\Dcal}^{\operatorname{trop}} \to \Ecal^{\operatorname{trop}}.
\]
The tropical curves on the left inherit a central alignment from the moduli space $\widetilde{\VZ}_{1,\alpha}(\PP^N|H,d)$, which orders a certain subset of the edge lengths. The polyhedral criterion for equidimensionality is that every cone of the source maps surjectively onto a cone of the target. The logarithmic modification $\widetilde\Ecal \to \Ecal$ is constructed by taking the minimal subdivison of $\Ecal^{\operatorname{trop}}$ such that this condition is satisfied.

By explicit but elementary polyhedral considerations, we are led to the following moduli theoretic description of the logarithmic modification $ \widetilde \Ecal$ such that the map $\widetilde \Dcal\to \widetilde \Ecal$ is flat.

We keep in mind the fibre product description of $\Ecal$. Construct a modification $\widetilde \Ecal$ from $\Ecal$ as follows:
\begin{enumerate}
\item Given a logarithmic map of type $\Delta$ with splitting nodes $q_1,\ldots,q_r$, formally glue together the $r$ vertices of the tropicalization which carry the splitting nodes, and declare the resulting vertex to be the root of the resulting tree. Impose the central alignment condition, where the special root takes on the r\^ole previously occupied by the core.
\item Replace the fibre product factor in $\Ecal$ associated to the vertex $\sqC_0$
\begin{equation*} \ol\Mcal^{\leftrightarrow}_{\Gamma_{\tiny{\sqC_0}}}(P|H_0+H_\infty) \end{equation*}
with the radially aligned rubber space:
\begin{equation*} \VZ^{\leftrightarrow}_{\Gamma_{\tiny{\sqC_0}}}(P|H_0+H_\infty).\end{equation*}
\end{enumerate}

This procedure induces a logarithmic modification $\widetilde\Ecal \to \Ecal$ which admits a modular description as a fibre product of genus zero centrally aligned maps coming from the vertices $\sqC_1,\ldots,\sqC_r$, and genus one aligned rubber maps satisfying fibre-direction factorisation from the vertex $\sqC_0$.

@@

We have produced a factorization
\begin{equation*}\widetilde\Dcal\to\widetilde \Ecal\to\Ecal,
\end{equation*}
where the first map is flat and the second map is a logarithmic modification. $\widetilde\Dcal$ is a compactification of a torus bundle over $\widetilde\Ecal$ whose strata can be read off from the tropicalization; in particular, we have the information of how (polynomials in) the classes of boundary strata push forward to $\widetilde\Ecal$. In fact, the generic fibre of $\widetilde\Dcal \to \widetilde\Ecal$ can be read off geometrically.

\begin{lemma}\label{lem:generic_proj_bundle}
 Generically on $\Ecal$, $\widetilde \Dcal\to\widetilde\Ecal$ is given by an explicit projective bundle.
\end{lemma}

@@

\subsection*{Step 4} Finally, consider a rubber space $\VZ^{\leftrightarrow}_{1,\alpha}(P|Y_0+Y_\infty,\beta)$ with insertion $\gamma$. Suppose inductively that we have computed all absolute and relative invariants with $d^\prime \leq d$ and all rubber invariants with $(d^\prime,n^\prime,m^\prime) < (d,n,m)$.

We first note the following important reduction: if there exists a relative marking $x_k$ such that all insertions are pulled back along $\fgt_{k}$, then we may apply the projection formula, together with the fact that
\begin{equation*} (\fgt_{k})_\st [\VZ^{\leftrightarrow}_{1,\alpha}(P|Y_0+Y_\infty,\beta)] = \alpha_k^2 \cdot [\VZ_{1,n+m-1}(Y,\beta)] \end{equation*}
to identify the rubber invariant with a multiple of a reduced invariant of $Y$, which has the same degree and hence is known recursively.

We will deal with the general case by reducing to the one above. Consider the moduli space
\begin{equation*}\label{step 4 recursion space} \VZ^{\leftrightarrow}_{1,\alpha\cup (0)}(P|Y_0+Y_\infty,\beta)\end{equation*}
obtained by introducing a marked point $y$ with no tangency. Let $x_1$ be a positive-tangency marking (such a marking always exists since $d \geq 2$) and let $\tilde\gamma$ be the insertion obtained from $\gamma$ by replacing $\ev_1$ and $\psi_1$ by $\ev_y$ and $\psi_y$, and then introducing $\fgt_{1}^\st$ everywhere.

We will make use of a recursion formula for rubber spaces analogous to the recursion formula for relative spaces used in Steps 1--3. Following \cite{EKatz}, there is a line bundle $L_y^{\not\in \operatorname{bot}}$ on $\VZ^{\leftrightarrow}_{1,\alpha\cup (0)}(P|Y_0+Y_\infty,\beta)$, together with a section $s_y^{\not\in\operatorname{bot}}$ whose vanishing locus consists of the locus $\Dcal(y)$ of rubber maps where $y$ is not mapped into the bottom level of the expanded target. As in \S \ref{Section Gathmann line bundle}, we can give a logarithmic interpretation of this: it corresponds to the piecewise-linear function on the tropical moduli space which associates, to every rubber tropical map, the distance between $\varphi(\sqC_y)$ and the leftmost vertex of the tropical target, where $\sqC_y$ is the vertex of the source curve containing the flag corresponding to $y$. Using this tropical description, we may easily calculate the vanshing orders of $s_y^{\not\in\operatorname{bot}}$ along the various components of $\Dcal(y)$, and show that $\cchern_1(L_y^{\not\in\operatorname{bot}}) = \Psi_0 - \ev_y^\st Y$ (see also \cite{EKatzLB}, where similar results in the non-reduced setting are obtained, using different methods). From this, we obtain a rubber recursion formula
\begin{equation}\label{step 4 recursion} (\Psi_0 - \ev_y^\st Y) \tilde\gamma \cap [\VZ^{\leftrightarrow}_{1,\alpha\cup(0)}(P|Y_0+Y_\infty,\beta)] = \tilde\gamma \cap [\Dcal(y)]\end{equation}

@@

\begin{equation*} \fgt_1^\st (-\alpha_2\psi_2 - \ev_y^\st Y)\tilde\gamma \cap [\VZ^{\leftrightarrow}_{1,\alpha\cup(0)}(P|Y_0+Y_\infty,\beta)] = \tilde\gamma\cap[\Dcal(y)]\end{equation*}

>>>>

\subsubsection{Type $B$}
Now suppose $\Delta$ is of type $B$. Note that in this case, it is impossible for the core to be contracted. Hence $\Ecal$ is disjoint from the blown-up locus, and the map $\Dcal \to \Ecal$ is an isomorphism. We thus obtain the following description, entirely in terms of genus zero data:
\begin{lemma} $\Dcal$ admits a finite and surjective splitting morphism:
\begin{equation*} \Dcal \xrightarrow{\rho} \left(\ol\Mcal_{0,\alpha^{(1)}\cup(m_1,m_2)}(\PP^N|H,d_1)\times\prod_{i=3}^r \ol\Mcal_{0,\alpha^{(i)}\cup(m_i)}(\PP^N|H,d_i)\right) \times_{H^r} \ol\Mcal^{\leftrightarrow}_{0,\alpha^{(0)}\cup(-m_1,\ldots,-m_r)}(\mathbb P|H_0+H_\infty,d_0).\end{equation*}\end{lemma}

@@

The combinatorics of the flattening can be described explicitly. Consider the map on tropicalizations
\[
\widetilde{\Dcal}^{\operatorname{trop}} \to \Ecal^{\operatorname{trop}}.
\]
The tropical curves on the left inherit a central alignment from the moduli space $\widetilde{\VZ}_{1,\alpha}(\PP^N|H,d)$, which orders a certain subset of the edge lengths. The polyhedral criterion for equidimensionality is that every cone of the source maps surjectively onto a cone of the target. There is a universal logarithmic modification $\widetilde\Ecal \to \Ecal$ such that $\widetilde\Dcal \to \widetilde\Ecal$ is flat, constructed by taking the minimal subdivison of $\Ecal^{\operatorname{trop}}$ such that cones map surjectively to cones.

This subdivision is best described using the formalism of extended tropicalisations \cite{}. The subdividing cones are obtained from cones in the star of $\tau$ under the limit $e_1=\cdots=e_r=\infty$. The cones in the star of $\tau$ are given by (in)equalities relating distances of pairs of vertices, and there are three cases to consider:
\begin{enumerate}
\item both vertices are components of the subgraph $\sqC \setminus \sqC_0$;
\item both vertices are components of the subgraph $\sqC_0$;
\item one vertex is a component of $\sqC_0$ and one is a component of $\sqC \setminus \sqC_0$.
\end{enumerate}
As we pass to the limit, the first relations impose a central alignment condition on the genus zero moduli, the second relations impose a radial alignment condition on the genus one moduli, and the third relations disappear entirely. We are therefore led to the following moduli theoretic description of the logarithmic modification $ \widetilde \Ecal$.

Keeping in mind the fibre product description of $\Ecal$, construct a modification $\widetilde \Ecal$ from $\Ecal$ as follows:
\begin{enumerate}
\item Given a logarithmic map of type $\Delta$ with splitting nodes $q_1,\ldots,q_r$, remove the genus one piece $\sqC_0$ and formally glue together the $r$ vertices of the tropicalization which carry the splitting nodes, and declare the resulting vertex to be the root of the resulting tree. Impose the central alignment condition, where the special root takes on the r\^ole previously occupied by the core.
\item Replace the fibre product factor in $\Ecal$ associated to the vertex $\sqC_0$
\begin{equation*} \ol\Mcal^{\leftrightarrow}_{\Gamma_{\tiny{\sqC_0}}}(\mathbb P|H_0+H_\infty) \end{equation*}
with the radially aligned rubber space:
\begin{equation*} \VZ^{\leftrightarrow}_{\Gamma_{\tiny{\sqC_0}}}(\mathbb P|H_0+H_\infty).\end{equation*}
(Here we are imposing fibre-direction factorisation in order to isolate the main component.)
\end{enumerate}
This procedure induces a logarithmic modification $\widetilde\Ecal \to \Ecal$ which admits a modular description as a fibre product of genus zero centrally aligned maps coming from the vertices $\sqC_1,\ldots,\sqC_r$, and genus one aligned rubber maps satisfying fibre-direction factorisation from the vertex $\sqC_0$.

@@

We have produced a factorization
\begin{equation*}\widetilde\Dcal\to\widetilde \Ecal\to\Ecal,
\end{equation*}
where the first map is flat and the second map is a logarithmic modification. $\widetilde\Dcal$ is a compactification of a torus bundle over $\widetilde\Ecal$ whose strata can be read off from the tropicalization; in particular, we have the information of how (polynomials in) the classes of boundary strata push forward to $\widetilde\Ecal$. In fact, we can say more and describe the generic fibre of $\widetilde\Dcal \to \widetilde\Ecal$:
\begin{lemma}\label{lem:generic_proj_bundle}
 Generically on $\Ecal$, $\widetilde \Dcal\to\widetilde\Ecal$ is given by an explicit projective bundle.
\end{lemma}

@@

\subsection*{Step 4} Finally, consider a rubber space $\VZ^{\leftrightarrow}_{1,\alpha}(\mathbb P|Y_0+Y_\infty,\beta)$ with insertion $\gamma$. Suppose inductively that we have computed all absolute and relative invariants with $d^\prime \leq d$ and all rubber invariants with $(d^\prime,n^\prime,m^\prime) < (d,n,m)$.

We first note the following important reduction: if there exists a relative marking $x_k$ such that all insertions are pulled back along $\fgt_{k}$, then we may apply the projection formula, together with the fact that
\begin{equation*} (\fgt_{k})_\st [\VZ^{\leftrightarrow}_{1,\alpha}(\mathbb P|Y_0+Y_\infty,\beta)] = \alpha_k^2 \cdot [\VZ_{1,n+m-1}(Y,\beta)] \end{equation*}
to identify the rubber invariant with a multiple of a reduced invariant of $Y$, which has the same degree and hence is known recursively.

We will deal with the general case by reducing to the one above. Consider the moduli space
\begin{equation*}\label{step 4 recursion space} \VZ^{\leftrightarrow}_{1,\alpha\cup (0)}(\mathbb P|Y_0+Y_\infty,\beta)\end{equation*}
obtained by introducing a marked point $y$ with no tangency. Let $x_1$ be a positive-tangency marking (such a marking always exists since $d \geq 2$) and let $\tilde\gamma$ be the insertion obtained from $\gamma$ by replacing $\ev_1$ and $\psi_1$ by $\ev_y$ and $\psi_y$, and then introducing $\fgt_{1}^\st$ everywhere.

We will make use of a recursion formula for rubber spaces analogous to the recursion formula for relative spaces used in Steps 1--3. Following \cite{EKatz}, there is a line bundle $L_y^{\not\in \operatorname{bot}}$ on $\VZ^{\leftrightarrow}_{1,\alpha\cup (0)}(\mathbb P|Y_0+Y_\infty,\beta)$, together with a section $s_y^{\not\in\operatorname{bot}}$ whose vanishing locus consists of the locus $\Dcal(y)$ of rubber maps where $y$ is not mapped into the bottom level of the expanded target. As in \S \ref{Section Gathmann line bundle}, we can give a logarithmic interpretation of this: it corresponds to the piecewise-linear function on the tropical moduli space which associates, to every rubber tropical map, the distance between $\varphi(\sqC_y)$ and the leftmost vertex of the tropical target, where $\sqC_y$ is the vertex of the source curve containing the flag corresponding to $y$. Using this tropical description, we may easily calculate the vanshing orders of $s_y^{\not\in\operatorname{bot}}$ along the various components of $\Dcal(y)$, and show that $\cchern_1(L_y^{\not\in\operatorname{bot}}) = \Psi_0 - \ev_y^\st Y$ (see also \cite{EKatzLB}, where similar results in the non-reduced setting are obtained, using different methods). From this, we obtain a rubber recursion formula
\begin{equation}\label{step 4 recursion} (\Psi_0 - \ev_y^\st Y) \tilde\gamma \cap [\VZ^{\leftrightarrow}_{1,\alpha\cup(0)}(\mathbb P|Y_0+Y_\infty,\beta)] = \tilde\gamma \cap [\Dcal(y)]\end{equation}

@@

\begin{equation*} \fgt_1^\st (-\alpha_2\psi_2 - \ev_y^\st Y)\tilde\gamma \cap [\VZ^{\leftrightarrow}_{1,\alpha\cup(0)}(\mathbb P|Y_0+Y_\infty,\beta)] = \tilde\gamma\cap[\Dcal(y)]\end{equation*}
